---
title: "Mass Spec Aging Data Analysis"
author: "Samuel Leung"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load and Preprocess data}
data <- read.csv("data/cardio_data_processed.csv")
head(data)
```

```{r Finding size of dataset}
nrow(data)
```

## Goal

-   To estimate the total number of people with cardiovascular disease: $\hat{t}_{\text{cardio}}$
-   To estimate the average blood pressure of the population: $\bar{u}_{\text{ap_hi}}, \hspace{1mm}\bar{u}_{\text{ap_lo}}$

## Sample Size Calculations:

We can't find the maximum variance from a continuous variable, we go through sample size calculations with the maximum variance possible from a binary variable: the total number of people with cardiovascular disease. We want the margin of error $\delta$ of our estimate to be within $\pm$ 1000 people, 19 times out of 20.

### Mathematical steps

Let $X$ be the binary random variable that models the number of people with cardiovascular disease, with mean $p$ and variance $p(1-p)$. Let $\bar y_S = \hat p$ be the estimate of the average number people with cardiovascular disease. of We want to estimate the total number of people with cardiovascular disease, $\hat t_{p,\text{cardio}} = N \bar y_s$. The variance of this estimator will be $var(\hat t_{p, \text{cardio}}) = N^2\hat p(1-\hat p)$. Going through our sample size calculations, we want to maximize variance, so we take $\hat p = 0.5$. Therefore, ignoring the finite population correction factor for now, we want to solve the following:

$$
\begin{align*}
1000 &\geq 1.96\sqrt{0.25 N^2 / n_0}\\
\implies \Big(\frac{1000}{1.96}\Big)^2 &\geq \frac{N^2}{4n_0}\\
\implies n_0 &\geq \Big(\frac{N}{2} \cdot \frac{1.96}{1000}\Big)^2\\
\end{align*}
$$ Converting this to code:

```{r Sample Size Calc}
N <- nrow(data)
target <- 1000
n_0 <- (1.96 * N / (2*target))^2
print(paste0("sample size ignoring FPC: ", n_0))
print(paste0("n_0 / N: ", n_0 / N))
```

Comparing the sample size calculated above with the total population size, we see that $\frac{n_0}{N} = 0.066 \geq 0.05$. So we cannot ignore the finite population correction factor in our sample size calculations. Factoring in the FPC, we would want to solve the following: $$n = \frac{n_0}{1+n_0/N}$$ Doing the calculations again with the FPC in mind with code:

```{r Sample Size with FPC}
n <- ceiling(n_0 / (1 + n_0 / N))
print(paste0("Sample Size with FPC: ", n))
```

So we use a sample size of 4194.

```{r Generate random sample}
# set seed for reproducable results
set.seed(124)

# generate random sample
srs <- data[sample(1:nrow(data), n), ]

head(srs)

# sanity check
print(nrow(srs))
```

```{r Calculate estimates}
N <- nrow(data)
fpc <- 1 - n / N

ap_hi.est <- round(mean(srs$ap_hi), 3)
ap_hi.var <- fpc*var(srs$ap_hi) / n
ap_hi.moe <- round(1.96*sqrt(ap_hi.var), 3)

ap_lo.est <- round(mean(srs$ap_lo), 3)
ap_lo.var <- fpc*var(srs$ap_lo) / n
ap_lo.moe <- round(1.96*sqrt(ap_lo.var), 3)

cardio.p <- mean(srs$cardio)
cardio.est <- round(N*cardio.p, 3)
cardio.var <- fpc*(cardio.p)*(1-cardio.p) * N
cardio.moe <- round(1.96*sqrt(cardio.var), 3)

print(paste0("Average ap_hi estimate: ", ap_hi.est, " +/- ", ap_hi.moe))
print(paste0("Average ap_lo estimate: ", ap_lo.est, " +/- ", ap_lo.moe))
print(paste0("Total cardio estimate: ", cardio.est, " +/- ", cardio.moe))
```

```{r TODO}
# sample size calcs
# from SRS: make estimates and calculate CI
```
